---
title: "Final Model Benchmark"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## My Final Model Benchmark

This report shows running benchmark on my final model. The benchmark comes from the [discussion forum of the course](https://www.coursera.org/learn/data-science-project/discussions/all/threads/1aWTuoCGEeakbhIiKPxV8w). The benchmark can be downloaded from [https://github.com/hfoffani/dsci-benchmark](https://github.com/hfoffani/dsci-benchmark)

## Running Benchmark On The Baseline
Running benchmark on the baseline. The baseline predicts the next word using the very common word such as "the", "on", "a".
```{r}
# load ngrams from saved files
source("load_ngram.R")

# load my predict nextword function
source("predict_nextword.R")

# load the benchmark function
source("benchmark.R")

# run the baseline benchmark 
benchmark(predict.baseline, sent.list = list('tweets' = tweets, 'blogs' = blogs)
          , ext.output = T)

```

## Running Benchmark On My Model

Running benchmark on my model

```{r}
benchmark(predict_nextwords, sent.list = list('tweets' = tweets, 'blogs' = blogs)
          , ext.output = T)

```


